Hierarchical Inference of the Intrinsic Ellipticity Distribution of Galaxies during Weak Lensing Analysis

### 2013-04-13 docs/ms.tex
Update ms.tex
update ms.tex

In galaxy--galaxy lensing, since the distortion is so weak, what people
usually do is to stack many lensing signals together. Then they can obtain the mean
shear map and infer the mean mass profile of single galaxy or galaxy
cluster. However with the Hierarchical Bayesian method we can avoid this brute averaging
of lensing signal. Instead, we can get the posterior distribution of mass profile for
each lensing system, even though it may have very large uncertainty. Then we
can calculate their statistical properties like galaxy-mass correlation
function via this posterior distribution.

@@

\section{method}

@@

and it's inverse formula 
\begin{equation}
\epsilon_{0}=\frac{\epsilon_{\ell}-g}{1-g^{*}\epsilon_{\ell}}
\end{equation}
from Schramm \& Kayser (1995), Seitz \& Schneider (1997), where $\epsilon$
is represented as a complex variable and g, $g^{*}$ are the reduced
shear and its complex conjugate, respectively. In weak lensing region we have 
\begin{equation}
g=\frac{\gamma}{1-k}\approx\gamma
\end{equation}
$\gamma$ is the shear. This is the assumption we use through out
the paper.$\epsilon$ is defined in terms of the major and minor axes and orientation a, b, $\phi$,
respectively, as
\begin{equation}
\epsilon=\frac{a-b}{a+b}e^{2i\phi}
\end{equation}
Due to the isotropicity of the Universe, for the unlensed ellipticity $\epsilon_{0}$ 
we expect $\phi$ has a uniform distribution
from 0 to $\pi$. Then we can expect 
\begin{equation}
<\epsilon_{\ell}>=g
\end{equation}
Therefore $<\epsilon_{\ell}>$ can be adopted as an estimator of g.

@@

Suppose we know the intrnsic ellipticity distribution P($\epsilon_{0}$)
based on some parasmeters $\alpha$. Then we can write down the likelihood
of reduced shear and $\alpha$:

@@

If there is measurement error on $\epsilon_{ob}$, we will have 
\begin{equation}
p(\epsilon_{ob}|\vec{g},\alpha)=\int f(\epsilon_{ob}|\epsilon_{\ell})\, p(\epsilon_{\ell}|\vec{g},\alpha)\,d\epsilon_{\ell}
\end{equation}
Here the prior distribtuion we use for $\vec{g}$ is a 2D flat distribution
in the region $|\vec{g}|<1$. Therefore $p(\vec{g)}=1$ if $|\vec{g}|<1$
and $p(\vec{g})=0$ if $|\vec{g}|>1$.

@@

One thing to note is that the likelihood of g in one patch only explicitly depends
on the shape parameters but not dependent to shear in other patches. Therefore, the
marginalized likelihood of shape parameters is :

@@

$\ell_{i}(\alpha|data_{i})$ is the marginalized likelihood function
calculated with data in the ith patch. From this formula we can see
how better estimation of $\alpha$ will be obtained if we infer more
sky patches simultaneously. With better knowledge on $\alpha$ we
can in turn have better estimates on reduced shear g.

\subsection{Model for Intrinsic Shape distribution}

The correctness of posterior distribution of g depends on the intrinsic
ellipticity distribution P($\epsilon_{0}|\alpha$). Due to the limitation
of human's knowledge we are not able to give an analytical form of
$P(\epsilon_{0}|\alpha)$. Therefore we proposed an exceedingly flexible
model for $P(\epsilon_{0}|\alpha)$, which does not requires further
assumption on $P(\epsilon_{0}|\alpha)$.

We divide the 2-D space r $\in$(0,1) into N bins. The probability
of each bin is $p_{i}=e^{\alpha_{i}}$. $\overrightarrow{\alpha}$
is a N-D parameter vector to be infered, simultaneously with g.

From practice we find that a prior smooth regularzation on $\alpha$
is needed to obtain a good estimation on shear g.

The prior distribution we used on $\alpha$ is : 
\begin{equation}
p(\vec{\alpha}|\xi)=exp(-\xi\sum(\alpha_{i}-\alpha_{i-1})^{2})
\end{equation}

@@

\subsection{Posterior samples}

@@

The probability distribution can be visualised by plotting points at each sample
position. An example in the x, y plane is shown for the posterior distribution of
$\vec{g}$ of 64 sky patches in Fig 1. There are 8 galaxies in each patch in the simulation.
The number density of samples is proportional to the probability density. The figure shows that the two parameters of $\vec{g}$
do not have strong correlation, since the shape of smaples is close to a circle.
We also shows the histogram of g1 in a single sky patch. Not surprisingly it is close to a Gaussian distribution.
If the probability distribution of parameters are perfect Gaussian, the most probable estimates and uncertainty 
is just the mean and standard deviation of the Gaussian distribution. However, in pratical, we find that the distribution 
of reduced shear $\vec{g}$ will  get closer to Gaussian, when larger number of galaxies in one patch  is used. When the number of galaxies in 
one patch is equal to or larger than 8, we found it's eligible to make such an approximation that to use sample mean and standard 
deviation as the most probable estimates and uncertainty. We'll justify this approximation with a bias test in the following
sub section. We also show a histgram of g1 in a random patch in fig 2.

@@

We calculated the value of $\hat{g}-g^{true}$ of 10,000 data. Each data set has 64 patches and 8$\times$64 galaxies, and
the true(input) P($\epsilon_{0}$) is Beta(2.8,2.8). The step function model is used with the prior prefator $\xi=3$.
The input true values of $\vec{g}_{1}$ and $\vec{g}_{2}$ of the 64$\times$10,000 sky patches are generated randomly from a uniform distribution from -0.1 to 0.1 .

@@

\subsection{Measurement Error}
In this section we  discuss how the shape measurement error will
affect the variance and bias of this Hierarchical Bayesian method.
In principle, if the likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$
is known the likelihood $p(\epsilon_{ob}|g,\alpha)$ can be calculated using the convolution as equation ().
However, due to the limitation of computational time we avoid to do this convolution but use $\epsilon_{ob}$
as an estimator of $\epsilon_{\ell}$.
The likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$ we use in the simulation is a 2D
Gaussian with mean equals $\epsilon_{\ell}$ and $\sigma$ equals 0.05 .
The result of bias test and mean square error test is shown in

\subsection{prefactor $\xi$}

>>>>

In galaxy--galaxy lensing, since the distortion is so weak, what people
usually do is to stack many lensing signals together. Then the mean
shear map and mean mass profile of single galaxy or galaxy
cluster can be inferred from this stacked lensing signal. However with the Hierarchical Bayesian method we can avoid this brute averaging
of lensing signal. Instead, we can get the posterior distribution of mass profile for
each lensing system, even though it may have very large uncertainty. Then we
can calculate their statistical properties like galaxy-mass correlation
function via this posterior distribution.

@@

\section{Method}

@@

and it's inverse formula 
\begin{equation}
\epsilon_{0}=\frac{\epsilon_{\ell}-g}{1-g^{*}\epsilon_{\ell}}
\end{equation}
from Schramm \& Kayser (1995), Seitz \& Schneider (1997), where $\epsilon$
is represented as a complex variable. It is defined in terms of the major and minor axes and orientation a, b, $\phi$,
respectively, as
\begin{equation}
\epsilon=\frac{a-b}{a+b}e^{2i\phi}
\end{equation}
$g$, $g^{*}$ are the reduced shear and its complex conjugate, respectively. In weak lensing region we have 
\begin{equation}
g=\frac{\gamma}{1-k}\approx\gamma
\end{equation}
$\gamma$ is the shear. This is the assumption we use through out
the paper.
Due to the isotropicity of the Universe, for the unlensed ellipticity $\epsilon_{0}$ 
we expect $\phi$ has a uniform distribution
from 0 to $\pi$. Then we can expect 
\begin{equation}
<\epsilon_{\ell}>=g
\end{equation}
Therefore $<\epsilon_{\ell}>$ can be adopted as an estimator of g.

@@

Suppose we know the intrnsic ellipticity distribution $P(\epsilon_{0}|\vec{\alpha})$
based on some parasmeters $\vec{\alpha}$. Then we can write down the joint likelihood
of reduced shear $g$ and shape parameters $\vec{\alpha}$:

@@

For a realistic unperfect galaxy shape measurement, there will be measurement errors on $\epsilon_{ob}$. Then we will have 
\begin{equation}
p(\epsilon_{ob}|\vec{g},\alpha)=\int f(\epsilon_{ob}|\epsilon_{\ell})\, p(\epsilon_{\ell}|\vec{g},\alpha)\,d\epsilon_{\ell}
\end{equation}
Here the prior distribtuion we use for $\vec{g}$ is a 2D flat distribution
in the region $|\vec{g}|<1$. Therefore $p(\vec{g)}=1$ if $|\vec{g}|<1$
and $p(\vec{g})=0$ if $|\vec{g}|>1$.

@@

One thing to note is that the likelihood of $g$ in one patch only explicitly depends
on the shape parameters but not dependent to reduced shear in other patches. Therefore, the
marginalized likelihood of shape parameters is :

@@

$\ell_{i}(\alpha|data_{i})$ is the marginalized likelihood function
calculated with data in the ith patch. From this formula we can see
how better inference of $\vec{\alpha}$ will be obtained if we infer numerous
sky patches simultaneously. With better knowledge on $\vec{\alpha}$ we
will in turn have better estimates on the reduced shear $g$.

\subsection{Model for Intrinsic Shape Distribution}

The correctness of posterior distribution of g depends on the intrinsic
ellipticity distribution P($\epsilon_{0}|\alpha$). Due to the limitation
of human's knowledge we will not be able to give an analytical form for
$P(\epsilon_{0}|\alpha)$. Therefore we proposed an exceedingly flexible
model for $P(\epsilon_{0}|\alpha)$, which does not requires further
assumption on $P(\epsilon_{0}|\alpha)$.

We divide the 2-D space r $\in$(0,1) into N bins. The probability
of each bin is $p_{i}=e^{\alpha_{i}}$. $\overrightarrow{\alpha}$
is a N dimension parameter to be infered, simultaneously with $g$.

From practice we find that a prior smooth regularizer on $\vec{\alpha}$
is needed to obtain a good estimation on $g$ and $\vec{\alpha}$.

The prior distribution we used on $\vec{\alpha}$ is : 
\begin{equation}
p(\vec{\alpha}|\xi)=exp(-\xi\sum(\alpha_{i}-\alpha_{i-1})^{2})
\end{equation}

@@

\subsection{Posterior Samples}

@@

The probability distribution can be visualised by plotting points at each sample
position. An example in the x, y plane is shown for the posterior distribution of
$\vec{g}$ of 64 sky patches in Figure 1. There are 8 galaxies in each patch in the simulation.
The number density of samples is proportional to the probability density. The figure shows that the two parameters of $\vec{g}$
do not have strong correlation, since the shape of smaples is close to a circle.
We also shows the histogram of g1 in a single sky patch in Figure 2. Not surprisingly it is close to a Gaussian distribution.
If the probability distribution of parameters are perfect Gaussian, the most probable estimates and uncertainty 
is just the mean and standard deviation of the Gaussian distribution. However, in pratical, we find that the distribution 
of reduced shear $\vec{g}$ will  get closer to Gaussian, when larger number of galaxies in one patch  is used. When the number of galaxies in 
one patch is equal to or larger than 8, we found it's eligible to make such an approximation that to use sample mean and standard 
deviation as the most probable estimates and uncertainty. We'll justify this approximation with a bias test in the following
sub section.

@@

We calculated the value of $\hat{g}-g^{true}$ of 10,000 data. Each data set has 64 patches and 8$\times$64 galaxies, and
the true(input) P($\epsilon_{0}$) is Beta(2.8,2.8) distribution. The step function model is used with the prior prefator $\xi=3$.
The input true values of $\vec{g}_{1}$ and $\vec{g}_{2}$ of the 64$\times$10,000 sky patches are generated randomly from a uniform distribution from -0.1 to 0.1 .

@@

\subsection{Measurement Error}
In this section we  discuss how the shape measurement error 
affect the variance and bias of this Hierarchical Bayesian method.
In principle, if the likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$
is known the likelihood $p(\epsilon_{ob}|g,\alpha)$ can be calculated using the convolution as equation ().
However, due to the limitation of computational time we avoid to do this convolution but use $\epsilon_{ob}$
as an estimator of $\epsilon_{\ell}$.
The likelihood of true lensed ellipticity $p(\epsilon_{ob}|\epsilon_{\ell})$ we use in the simulation is a 2D
Gaussian with mean equals $\epsilon_{\ell}$ and $\sigma$ equals 0.05 .
The result of bias test and mean square error test is shown in

\subsection{Prefactor $\xi$}
